A performance comparison of Dask and Apache Spark for data-intensive neuroimaging pipelines

### 2019-07-30 latex/paper.tex
typos

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines has emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} BigBrain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on a 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both engines perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either engine is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data engines, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

@@

The next section details the design of our benchmarking experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the BigBrain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The other sections present
our results, discussion and conclusions.

@@

As data transfers in Big Data workflows are an important source of performance
bottlenecks, Spark incorporates the concepts of data locality and in-memory 
computing. Data locality, popularized by MapReduce~\cite{dean2008mapreduce},
schedules
tasks as close as possible to where the data is stored. In-memory computing
ensures data is maintained in memory whenever possible, as writing large 
amounts data to slower storage devices may be costly. To reduce any unnecessary
communication and computation, Spark also included lazy evaluation, which
builds the entire task graph prior to execution to determine what needs to be
computed.

@@

\subsubsection{Dask} Dask is a Python-based Big Data engine that is becoming
increasingly popular in the scientific Python ecosystem. Like Spark, Dask avoids data transfers and needless computation and communication through
in-memory computing, data locality, and lazy evaluation. Dask workflows can 
further reduce data transfer costs by leveraging multithreading whenever 
communication is not bounded by Python's GIL. Dask relies on data 
lineage to achieve fault tolerance, however, unlike Spark, does not require
operations to be coarse grained. Furthermore, Dask is lightweight and modular,
allowing users to only install the components they require.

Dask has five main data structures:
\href{https://docs.dask.org/en/latest/array.html}{Array}, \href{https://docs.dask.org/en/latest/bag.html}{Bag},
\href{https://docs.dask.org/en/latest/dataframe.html}{DataFrame},
 \href{https://docs.dask.org/en/latest/delayed.html}{Delayed}, 
 and \href{https://docs.dask.org/en/latest/futures.html}{Futures}. A Dask
Array is used for the
processing of large arrays. It provides a distributed clone of the popular NumPy
library. Similar to Spark's RDD, a Dask
Bag is a parallel
collection of Python objects. It offers a programming abstraction similar to the
\href{https://toolz.readthedocs.io/en/latest/}{PyToolz library}. A Dask
Dataframe is a parallel
composition of
\href{http://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html}{Pandas Dataframes},
 used to process a large amount of tabular data.
Dask Delayed supports
arbitrary tasks
that do not fit in the Array, DataFrame or Bag APIs. Finally,
Dask Futures are similar to
Delayed in that they support arbitrary tasks, but they operate in real-time
rather than lazily. With the exception of Dask Bag, all APIs, by default, use
the local multithreaded scheduler. Dask Bag, instead, relies on the local
multiprocessing scheduler.  All Dask data structures, with the exception of
Dask Array and Dask DataFrame, were used in our experiments.

@@

%% Incrementation
\subsubsection{Incrementation}
We used an adaptation of the image incrementation pipeline used
in~\cite{hayot2019performance} (see Algorithm~\ref{alg:incrementation}).
The application reads blocks of the BigBrain image from the shared file
system, increments the intensity value of each voxel by 1 to avoid caching
effects, sleeps for a configurable amount of time to emulate more
compute-intensive processing, repeats this process for a specified number of
iterations, and
finally writes the result as a NIfTI image back to the shared file system.
This application allows us to study the behaviour of the engines when all
inputs are processed independently, in a map-only scenario (see
Figure~\ref{fig:tg-inc}). This mimics the behaviour of analyzing multiple
independent subjects in parallel.

@@

 Our second application calculates the histogram of the BigBrain image (see
 Algorithm~\ref{alg:histogram}). It reads the image blocks from the shared
 file system, calculates intensity frequencies, aggregates the frequencies
 across the blocks, and finally writes the resulting histogram to the
 shared file system as a single \SI{766}{\kilo\byte} file. It follows the
 map-reduce paradigm, in which the final result is obtained from all the
 input blocks (see Fig.~\ref{fig:tg-histo}). This application requires data
 shuffling, incurring additional performance costs compared to
 Algorithm~\ref{alg:incrementation}. The total amount of shuffled data is,
 however, limited to \SI{2.62}{\mega\byte} per blocks as it only consists
 of image histograms. Two implementation are studied: a pure Python one,
 where frequencies are computed through Python dictionary manipulations, and one based
 on the \href{https://github.com/numpy}{NumPy} library, that implements array computations in C.

@@

While Incrementation and Histogram were implemented directly in Python, the 
BIDS App example requires an external set of command-line tools distributed as a Docker container image
(bids/base\_fsl on DockerHub). We converted the Docker image to a Singularity image for use in 
an HPC environment, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}.
The image was preloaded on the NFS and Singularity version 3.2.1\-1.el7 was installed
on all instances.

@@

On Figure~\ref{fig:inc_tt_worker}, we also note that Spark's overhead is
slightly lower than Dask's, particularly when the number of workers
increase. Moreover, Delayed and Futures have a higher overhead than Bag.
However, overhead differences are compensated by an increase in data
transfer time, as a reduced overhead increases the concurrency between data
transfers and the application saturates the 10Gbits/s network bandwidth.

@@

From Figure~\ref{fig:histo_tt_worker}, the total time spent in each function is
shown. The compute time is significantly larger for Dask than for Spark. This is
presumably due to Python's GIL preventing Dask to parallelize the computation
over
multiple threads. Overall, the I/O and overhead times are comparable for all engines.

@@

Figure~\ref{fig:histo_np_tt_block} shows the total time spent in each function.
Once more, since Delayed is thread-based its overhead at 30 blocks is seems much
higher that it really is. Considering the overhead error and that at 30 blocks
the workers only use half their core, the total time is substantially the same
for all block size. This is due to lower block sizes having a lower data transfer time,
however, it balances out by the resulting increase in overhead.

@@

In Figure~\ref{fig:histo_np_gantt}, read tasks are initially similar for all
engines. However, when computation starts the subsequent read tasks are much
shorter for Delayed. This is due the desynchronization of the task reducing the
data transfer bottleneck.

@@

%%% BIDS EXAMPLE %%%
\subsection{BidsApp example: Number of workers}
Figure~\ref{fig:bids_ms_worker} shows the makespan of the application when
varying the amount of workers. Overall, there is no substantial difference
between the engines. The makespan scales well with the increase in workers,
however, the scaling reduces slightly when reaching 8 workers. This is due
to an increase in I/O time caused by parallel accesses to 
the shared file system.

@@

Overall, our experiments did not show any substantial performance
difference between Spark and Dask. In contradiction with the experiments
of~\cite{Mehta:17}, we did not observe consequential overhead differences
due to serialization or startup costs. This apparent contradiction is most
likely coming from differences in applications, engine parametrizations, or
infrastructure characteristics. In our case, potential overhead differences were
compensated by reverse trends in data transfer time. Since our focus are
Big Data applications, this leads us to conclude that engines are
equivalent.

@@

The observed almost exact compensation between differences in engine
overheads and in data transfer times (see for instance
Figure~\ref{fig:inc_tt_worker}) is not surprising. Indeed, when the transfer
bandwidth is saturated, and in absence of service thrashing,
desynchronizing file transfers usually does not reduce the makespan of data
transfers. To take an extreme example, the makespan of $n$ concurrent file
transfers of size $F$ on a system of bandwidth $B$ would be $nF/B$, as $n$
transfers would share the bandwidth, which equals to the time required
to transfer the $n$ files sequentially. In other words, saturated bandwidths
give room for extra engine overhead as long as it desynchronizes data
transfers. This could be an interesting idea to explore in data transfer optimization.

@@

\section{Conclusion}
We presented a comparison of two Big Data engines, Apache Spark and Dask.
We studied the engines on three data-intensive neuroimaging applications
representative of common use cases. Overall, our results show no substantial
performance difference between the engines. Interestingly, differences in
engine overheads do not impact performance due to their impact on data
transfer times: higher overheads are almost exactly compensated by a lower
transfer times when data transfers saturate the bandwidth. These results
suggest that future research should focus on strategies to reduce the
impact of data transfers on applications.

>>>>

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines have emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} BigBrain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on an 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both engines perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either engine is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality, and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data engines, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

@@

The next section details the design of our benchmarking experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the BigBrain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The other sections present
our results, discussion, and conclusions.

@@

As data transfers in Big Data workflows are an important source of performance
bottlenecks, Spark incorporates the concepts of data locality and in-memory
computing. Data locality, popularized by MapReduce~\cite{dean2008mapreduce},
schedules tasks as close as possible to where the data is stored. In-memory computing
ensures data is maintained in memory whenever possible, as writing large amounts of
data to slower storage devices may be costly. To reduce any unnecessary communication
and computation, Spark also included lazy evaluation, which builds the entire task
graph prior to execution to determine what needs to be computed.

@@

\subsubsection{Dask} Dask is a Python-based Big Data engine that is becoming
increasingly popular in the scientific Python ecosystem. Like Spark, Dask avoids data transfers and needless computation and communication through
in-memory computing, data locality, and lazy evaluation. Dask workflows can 
further reduce data transfer costs by leveraging multithreading whenever 
communication is not bounded by Python's GIL. Dask relies on data 
lineage to achieve fault tolerance, however, unlike Spark, it does not require
operations to be coarse-grained. Furthermore, Dask is lightweight and modular,
allowing users to only install the components they require.

Dask has five main data structures:
\href{https://docs.dask.org/en/latest/array.html}{Array}, \href{https://docs.dask.org/en/latest/bag.html}{Bag},
\href{https://docs.dask.org/en/latest/dataframe.html}{DataFrame},
 \href{https://docs.dask.org/en/latest/delayed.html}{Delayed}, 
 and \href{https://docs.dask.org/en/latest/futures.html}{Futures}. A Dask
Array is used for the
processing of large arrays. It provides a distributed clone of the popular NumPy
library. Similar to Spark's RDD, a Dask
Bag is a parallel
collection of Python objects. It offers a programming abstraction similar to the
\href{https://toolz.readthedocs.io/en/latest/}{PyToolz library}. A Dask
Dataframe is a parallel
composition of
\href{http://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html}{Pandas Dataframes},
 used to process a large amount of tabular data.
Dask Delayed supports
arbitrary tasks
that do not fit in the Array, DataFrame or Bag APIs. Finally,
Dask Futures are similar to
Delayed in that they support arbitrary tasks, but they operate in real-time
rather than lazily. Except for Dask Bag, all APIs, by default, use
the local multithreaded scheduler. Dask Bag, instead, relies on the local
multiprocessing scheduler.  All Dask data structures, except for
Dask Array and Dask DataFrame, were used in our experiments.

@@

%% Incrementation
\subsubsection{Incrementation}
We used an adaptation of the image incrementation pipeline used
in~\cite{hayot2019performance} (see Algorithm~\ref{alg:incrementation}).
The application reads blocks of the BigBrain image from the shared file
system, increments the intensity value of each voxel by 1 to avoid caching
effects, sleeps for a configurable amount of time to emulate more
compute-intensive processing, repeats this process for a specified number of
iterations, and
finally writes the result as a NIfTI image back to the shared file system.
This application allows us to study the behavior of the engines when all
inputs are processed independently, in a map-only scenario (see
Figure~\ref{fig:tg-inc}). This mimics the behavior of analyzing multiple
independent subjects in parallel.

@@

 Our second application calculates the histogram of the BigBrain image (see
 Algorithm~\ref{alg:histogram}). It reads the image blocks from the shared
 file system, calculates intensity frequencies, aggregates the frequencies
 across the blocks, and finally writes the resulting histogram to the
 shared file system as a single \SI{766}{\kilo\byte} file. It follows the
 map-reduce paradigm, in which the final result is obtained from all the
 input blocks (see Fig.~\ref{fig:tg-histo}). This application requires data
 shuffling, incurring additional performance costs compared to
 Algorithm~\ref{alg:incrementation}. The total amount of shuffled data is,
 however, limited to \SI{2.62}{\mega\byte} per block as it only consists
 of image histograms. Two implementations are studied: a pure Python one,
 where frequencies are computed through Python dictionary manipulations, and one based
 on the \href{https://github.com/numpy}{NumPy} library, that implements array computations in C.

@@

While Incrementation and Histogram were implemented directly in Python, the 
BIDS App example requires an external set of command-line tools distributed as a Docker container image
(bids/base\_fsl on DockerHub). We converted the Docker image to a Singularity image for use in 
an HPC environment, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}.
The image was preloaded on the NFS and Singularity version 3.2.1\-1.el7 was installed
for all instances.

@@

On Figure~\ref{fig:inc_tt_worker}, we also note that Spark's overhead is
slightly lower than Dask's, particularly when the number of workers
increases. Moreover, Delayed and Futures have a higher overhead than Bag.
However, overhead differences are compensated by an increase in data
transfer time, as a reduced overhead increases the concurrency between data
transfers and the application saturates the 10Gbits/s network bandwidth.

@@

From Figure~\ref{fig:histo_tt_worker}, the total time spent in each function is
shown. The computing time is significantly larger for Dask than for Spark. This is
presumably due to Python's GIL preventing Dask to parallelize the computation
over
multiple threads. Overall, the I/O and overhead times are comparable for all engines.

@@

Figure~\ref{fig:histo_np_tt_block} shows the total time spent in each function.
Once more, since Delayed is thread-based its overhead at 30 blocks is much
higher. Considering the overhead error and that at 30 blocks
the workers only use half their core, the total time is substantially the same
for all block size. This is due to lower block sizes having a lower data transfer time,
however, it balances out by the resulting increase in overhead.

@@

In Figure~\ref{fig:histo_np_gantt}, read tasks are initially similar for all
engines. However, when computation starts the subsequent read tasks are much
shorter for Delayed. This is due to the desynchronization of the task reducing the
data transfer bottleneck.

@@

%%% BIDS EXAMPLE %%%
\subsection{BidsApp example: Number of workers}
Figure~\ref{fig:bids_ms_worker} shows the makespan of the application when
varying the number of workers. Overall, there is no substantial difference
between the engines. The makespan scales well with the increase in workers,
however, the scaling reduces slightly when reaching 8 workers. This is due
to an increase in I/O time caused by parallel accesses to 
the shared file system.

@@

Overall, our experiments did not show any substantial performance
difference between Spark and Dask. In contradiction to the experiments
of~\cite{Mehta:17}, we did not observe consequential overhead differences
due to serialization or startup costs. This apparent contradiction is most
likely coming from differences in applications, engine parametrizations, or
infrastructure characteristics. In our case, potential overhead differences were
compensated by reverse trends in data transfer time. Since our focus is
Big Data applications, this leads us to conclude that engines are
equivalent.

@@

The observed almost exact compensation between differences in engine
overheads and in data transfer times (see for instance
Figure~\ref{fig:inc_tt_worker}) is not surprising. Indeed, when the transfer
bandwidth is saturated, and in the absence of service thrashing,
desynchronizing file transfers usually does not reduce the makespan of data
transfers. To take an extreme example, the makespan of $n$ concurrent file
transfers of size $F$ on a system of bandwidth $B$ would be $nF/B$, as $n$
transfers would share the bandwidth, which equals to the time required
to transfer the $n$ files sequentially. In other words, saturated bandwidths
give room for extra engine overhead as long as it desynchronizes data
transfers. This could be an interesting idea to explore in data transfer optimization.

@@

\section{Conclusion}
We presented a comparison of two Big Data engines, Apache Spark and Dask.
We studied the engines on three data-intensive neuroimaging applications
representative of common use cases. Overall, our results show no substantial
performance difference between the engines. Interestingly, differences in
engine overheads do not impact performance due to their impact on data
transfer time: higher overheads are almost exactly compensated by a lower
transfer time when data transfers saturate the bandwidth. These results
suggest that future research should focus on strategies to reduce the
impact of data transfers on applications.
